### 2016-01-27 Paper_v4.tex
Figs Updated

This particle was then updated according to the following rule
\begin{eqnarray}
	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
\end{eqnarray}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$. $\sigma$ is the amplitude of perturbation given by following equation:
\begin{eqnarray}
	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{eqnarray}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
	\mathcal{P}_{j}&=&{1}-\log(j/ (\({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
%where $\beta$ is the perturbation frequency probability modulator. 
Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$. The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS. 
The fraction of evaluations $\mathcal{FR}$  within the swarm phase and DDS phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations. This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.

\subsection*{Performance of DOPS on a model of the human coagulation cascade.}
We compared the performance of DOPS on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS). Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig. \ref{fig-coagulation-network}). The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}. This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets. The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.   Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.

 To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants. The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}. We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment. DdPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3). Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error. Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly. Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms. Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations. However the drop in error remains constant for DDS after 2000 function evaluations. Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2). The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval. Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).

>>>>

This particle was then updated according to the following rule
\begin{eqnarray}
	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
\end{eqnarray}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$. $\sigma$ is the amplitude of perturbation given by following equation:
\begin{eqnarray}
	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{eqnarray}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
	\mathcal{P}_{j}&=&{1}-\log(j/ (({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
%where $\beta$ is the perturbation frequency probability modulator. 
Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$. The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS. 
The fraction of evaluations $\mathcal{FR}$  within the swarm phase and DDS phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations. This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.

\subsection*{Performance of DOPS on a model of the human coagulation cascade.}
We compared the performance of DOPS on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), and dynamically dimensioned search (DDS). Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig. \ref{fig-coagulation-network}). The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}. This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets.

%Added a little more description here
The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.  For example when coagulation is initiated through tissue factor pathway, tissue factor and activated factor VIIa (FVIIa) form a complex that activates factors X (fX) and IX (fIX). Activated factor X (fXa) thereafter activates downstream factors VIII and IX. The initial activation leads to the production of picomolar amounts of thrombin (fIIa) which activates platelets and amplifies its own production through the formation of a prothrombinase complex (FXa-FVa) on the surface of the activated platelets. Thrombin also  downregulates its own production by forming a complex with thrombomodulin which then activates protein C (PC). PC inhibits the formation of prothrombinase complex. In addition, Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner. Antithrombin III (ATIII)  inhibits all proteases. surface protein thrombomodulin (TM). Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.

To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants. The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}. This choice of using a linear combination of two different error functions was motivated by poor validation results while using a single error function.  We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment. DOPS exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3).

 Within the first 1000 function evaluations  of DOPS there is a very rapid drop in error. Approximately between 500 -1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varies from trial to trial since the switch is based on the error from the swarm phase). Overall at the end of 4000 function evaluations DOPS minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms. Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 4). The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval. Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 5).
